Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/evaluation.tex
up to related
11 pages exactly!

\subsection{Trace Complexity}
\label{sec:trace-complexity}
For each of the ill-typed programs for which we could find a witness, we
measure the complexity of the generated trace according to the following
metrics:
%
\begin{enumerate}
\item The number of single-step reductions along the path from the
  witness to the stuck term. This can be thought of as a worst-case
  complexity, \ie ``How big is the fully-expanded trace?''
\item The number forward (or backward) jumps that can be taken along the
  path from the witness to the stuck term\footnote{This is not quite the same as
  a stack-trace as it shows \emph{all} function calls, including those
  that returned successfully.}.
  % These abstract many details of the computation that are often
  % uninteresting or irrelevant to the explanation.
% \item \ES{others?}
\end{enumerate}
%
\begin{figure}[t]
\centering
% \includegraphics[width=\linewidth]{trace_size.pdf}
% \begin{minipage}{0.49\linewidth}
\includegraphics[width=\linewidth]{trace_size_step.pdf}
% \end{minipage}
% \begin{minipage}{0.49\linewidth}
\includegraphics[width=\linewidth]{trace_size_jump.pdf}
% \end{minipage}
\caption{Complexity of the generated traces. At least 90\% of the traces
  have a jump complexity of less than 10, with an average of 3.}
\label{fig:results-complexity}
\end{figure}
%

@@

\paragraph{Infinite Types}
Our implementation does check for infinite types inside \forcesym, but
there are some degenerate cases where it is unable to detect
them. Consider, for example, the following buggy version of @replicate@,
which is supposed to produce a list containing @n@ copies of @x@.
%
\begin{code}
  let rec replicate n x =
    if n <= 0 then
      []
    else
      replicate (n-1) [x]
\end{code}
%
This code produces a nested list (with @n@ levels of nesting) containing
a single copy of @x@. \ocaml detects a cyclic \hbox{@'a = 'a list@}
constraint in the recursive call and throws a type error, whereas
\nanomaly happily recurses @n@ times to produce the nested list.
Strictly speaking, this function itself cannot ``go wrong'', the
program would not get stuck until a \emph{client} attempted to use the
result expecting a flat list. But this is not very satisfying as
@replicate@ is clearly to blame. Furthermore, in our experience,
infinite-type errors are some of the more difficult ones to debug (and
to explain to novices), so better support for this scenario would be
useful.

>>>>

\subsection{Trace Complexity}
\label{sec:trace-complexity}
For each of the ill-typed programs for which we could find a witness, we
measure the complexity of the generated trace according to the following
metrics:
%
\begin{enumerate}
\item The number of single-step reductions along the path from the
  witness to the stuck term.
  % This can be thought of as a worst-case
  % complexity, \ie ``How big is the fully-expanded trace?''
\item The number forward (or backward) jumps that can be taken along the
  path from the witness to the stuck term\footnote{This is not quite the same as
  a stack-trace as it shows \emph{all} function calls, including those
  that returned successfully.}.
  % These abstract many details of the computation that are often
  % uninteresting or irrelevant to the explanation.
% \item \ES{others?}
\end{enumerate}
%
\begin{figure}[t]
\centering
% \includegraphics[width=\linewidth]{trace_size.pdf}
% \begin{minipage}{0.49\linewidth}
\includegraphics[width=\linewidth]{trace_size_step.pdf}
% \end{minipage}
% \begin{minipage}{0.49\linewidth}
\includegraphics[width=\linewidth]{trace_size_jump.pdf}
% \end{minipage}
\caption{Complexity of the generated traces. At least 90\% of the traces
  have a jump complexity of less than 10, with an average of 3.}
\label{fig:results-complexity}
\end{figure}
%

@@

\paragraph{Infinite Types}
Our implementation does check for infinite types inside \forcesym, but
there are some degenerate cases where it is unable to detect
them. Consider, for example, the following buggy version of @replicate@.
% ,
% which is supposed to produce a list containing @n@ copies of @x@.
%
\begin{code}
  let rec replicate n x =
    if n <= 0 then [] else replicate (n-1) [x]
\end{code}
%
This code produces a nested list (with @n@ levels of nesting) containing
a single copy of @x@, instead of a list with @n@ copies of @x@. \ocaml
detects a cyclic \hbox{@'a = 'a list@} constraint in the recursive call
and throws a type error, whereas \nanomaly happily % recurses @n@ times to
produces the nested list.  Strictly speaking, this function itself cannot
``go wrong'', the program would not get stuck until a \emph{client}
attempted to use the result expecting a flat list. But this is not very
satisfying as @replicate@ is clearly to blame. Furthermore, in our
experience, infinite-type errors are often difficult to %some of the more difficult ones to
debug (and to explain to novices), so better support for this scenario
would be useful.

### 2015-11-20 paper/pldi16-submission/related.tex
11 pages exactly!

Lerner \etal~\cite{lerner_searching_2007} attempt to localize the type
error and suggest a fix by replacing expressions (or removing them
entirely) with alternatives based on the surrounding program context.
%
Chen and Erwig~\cite{chen_counter-factual_2014} use a variational type
system represent the possibility that the inferred type of an expression
may be incorrect, because the expression may be the source of the
error. It then attempts to deduce the error source by searching for an
expression whose type can be changed such that type inference would
succeed. In contrast to Seminal, which searches for changes at the
value-level, by searching at the type level the search is complete due
the finite universe of types in a program.
%
Neubauer and Thiemann~\cite{neubauer_discriminative_2003} present a
decidable type system based on discriminative sum types, in which all
terms are typeable and type derivations contain all type errors in a
program. They then use the typing derivation to slice out the parts of
the expression related to each error.
%
Zhang and Myers~\cite{zhang_toward_2014} present an algorithm for
identifying the most likely culprit in a system of unsatisfiable
constraints (\eg type equalities), based on Bayesian reasoning.
%
Pavlinovic \etal~\cite{pavlinovic_finding_2014} translate the error
localization problem to a MaxSMT optimization problem, using
compiler-provided weights to rank the possible sources.
% \item \cite{chen_error-tolerant_2012}
% \item \cite{okeefe_type_1992}
% \item \cite{gomard_partial_1990}
% \item \cite{thatte_type_1988}

@@

\paragraph{Running Ill-Typed Programs}
\label{sec:running-ill-typed}
Vytiniotis \etal~\cite{vytiniotis_equality_2012} extend the Haskell
compiler GHC to support compiling ill-typed programs, but their intent
is rather different from ours. Their goal was to allow programmers to
incrementally test refactorings, which often cause type errors in
distant pieces of code. They replace any expression that fails to type
check with a \emph{runtime} error, but do not perform any type checking
at runtime.

@@

Instead of enumerating values, which may trigger the same path through
the program, one might enumerate paths. 
%
Dynamic-symbolic execution~\cite{godefroid_dart:_2005,cadar_klee:_2008,tillmann_pex_2008}
%
enables this by symbolic execution (to track which path a given input
vector triggers) with concrete execution (to ensure test failures are
not spurious). The system collects a path condition during execution,
which tracks symbolically what conditions must be met to trigger the
current path. Upon successfully completing a test run, it negates the
path condition and queries a solver for another set of inputs that
satisfy the negated path condition, \ie inputs that will not trigger the
same path. Thus, dynamic-symbolic execution can prune the search space
much faster than techniques based on enumerating values, but is limited
by the expressiveness of the underlying solver. Our operational
semantics is amenable to dynamic-symbolic execution, one would just need
to collect the path condition and replace our implementation of \gensym
by a call to the solver. We chose to use lazy, random generation instead
because it does not incur the overhead of an external solver, and
produces high coverage for our domain of novice programs.

@@

\subsection{Program Exploration}
Perera \etal~\cite{perera_functional_2012} present a tracing semantics
for functional programs that tags values with their provenance, enabling
a form of backwards program slicing from a final value to the sequence
of reductions that produced it. Notably, they allow the user to supply a
\emph{partial value} -- a value containing holes -- and in turn present
a partial slice, containing only those steps that affected the specified
components of the partial value. This system is designed to answer
questions of the form ``Where did this value come from?'' and thus is
focused on backward exploration. In contrast, our trace visualization
supports forward \emph{and} backward exploration, though our backward
steps are more limited than Perera's. Specifically, we do not support
selecting a value and inserting the intermediate terms that preceded it
while ignoring unrelated computation steps; this would be interesting
future work.

\ES{todo: more on program exploration}

>>>>

Lerner \etal~\cite{lerner_searching_2007} attempt to localize the type
error and suggest a fix by replacing expressions (or removing them
entirely) with alternatives based on the surrounding program context.
%
Chen and Erwig~\cite{chen_counter-factual_2014} use a variational type
system to allow for the possibility of changing an expression's type,
and search for an expression whose type can be changed such that type
inference would succeed.
%
% It then attempts to deduce the error source by searching for an
% expression whose type can be changed such that type inference would
% succeed.
%
In contrast to Lerner, who searches for changes at the
value-level, by searching at the type level Chen's search is complete due
the finite universe of types in a program.
%
Neubauer and Thiemann~\cite{neubauer_discriminative_2003} present a
decidable type system based on discriminative sum types, in which all
terms are typeable and type derivations contain all type errors in a
program. They then use the typing derivation to slice out the parts of
the expression related to each error.
%
Zhang and Myers~\cite{zhang_toward_2014} present an algorithm for
identifying the most likely culprit in a system of unsatisfiable
constraints (\eg type equalities), based on Bayesian reasoning.
%
Pavlinovic \etal~\cite{pavlinovic_finding_2014} translate the %error
localization problem to a MaxSMT optimization problem, using
compiler-provided weights to rank the possible sources.
% \item \cite{chen_error-tolerant_2012}
% \item \cite{okeefe_type_1992}
% \item \cite{gomard_partial_1990}
% \item \cite{thatte_type_1988}

@@

\paragraph{Running Ill-Typed Programs}
\label{sec:running-ill-typed}
Vytiniotis \etal~\cite{vytiniotis_equality_2012} extend the Haskell
compiler GHC to support compiling ill-typed programs, but their intent
is rather different from ours. Their goal was to allow programmers to
incrementally test refactorings, which often cause type errors in
distant functions. They replace any expression that fails to type
check with a \emph{runtime} error, but do not check types
at runtime.

@@

Instead of enumerating values, which may trigger the same path through
the program, one might enumerate paths. 
%
Dynamic-symbolic execution~\cite{godefroid_dart:_2005,cadar_klee:_2008,tillmann_pex_2008}
%
combines symbolic execution (to track which path a given input
triggers) with concrete execution (to ensure failures are
not spurious). The system collects a path condition during execution,
which tracks symbolically what conditions must be met to trigger the
current path. Upon successfully completing a test run, it negates the
path condition and queries a solver for another set of inputs that
satisfy the negated path condition, \ie inputs that will not trigger the
same path. Thus, it can prune the search space
much faster than techniques based on enumerating values, but is limited
by the expressiveness of the underlying solver. Our operational
semantics is amenable to dynamic-symbolic execution, one would just need
to collect the path condition and replace our implementation of \gensym
by a call to the solver. We chose to use lazy, random generation instead
because it is efficient, and % does not incur the overhead of an external solver, and
produces high coverage for our domain of novice programs.

@@

\subsection{Program Exploration}
Perera \etal~\cite{perera_functional_2012} present a tracing semantics
for functional programs that tags values with their provenance, enabling
a form of backwards program slicing from a final value to the sequence
of reductions that produced it. Notably, they allow the user to supply a
\emph{partial value} -- containing holes -- and present a partial slice,
containing only those steps that affected the the partial value. This
system is designed to answer questions of the form ``Where did this
value come from?'' and thus is focused on backward exploration. In
contrast, our trace visualization supports forward \emph{and} backward
exploration, though our backward steps are more limited than
Perera's. Specifically, we do not support selecting a value and
inserting the intermediate terms that preceded it while ignoring
unrelated computation steps; this would be interesting future work.

% \ES{todo: more on program exploration}
